---
title: "Response Time Analysis"
output: html_document
date: "2025-12-07"
---


## Introduction

In this file, you find part of our results for research question 1 (specifically in regards to response time and exercise). We start this file with loading data (the same as all other files) and then delve into our main findings. Readers interested in verifying the main result in RQ1 should search for the string "Research Question 1". 


## Packages

```{r}
# import packages
library(lme4)     # mixed models
library(dplyr)    # dataframe manipulation
library(readr)    # reading in the data
library(emmeans)  # likelihood ratio testing
library(purrr)    # for map()
library(lattice)  # for xyplot()
library(TAF)      # setting up output file directory structure during model fitting
library(moments)  # skewness for selecting between additive and multiplicative models of response times
library(performance)

library(ggplot2)


#If you get an error, you likely need to install the packages using the "install.package('insert-package-name-here')" command
```


## Notebook configurable settings

```{r}
# name the directories to read data from and write output to
input_version = 2
output_version = 'programming-v2'


```

the following code is to make an output directory
```{r}

outputdir = sprintf("%s/../candidate-models/regression-models-%s",getwd(), output_version)
mkdir(outputdir) #only uncomment this if you don't already have an output directory

```




## Loading Data

For consistency, we output the number of rows in our data which represent trials or items -- each row represents an instance where a participant answered a stimulus question.
```{r}

input_file = sprintf("%s/../preprocessed-data/full-data-version-%d/programming-full-data.csv", getwd(), input_version)
data <- read.csv(input_file)


nrow(data)

```


## Data Management

In this step, we want to calculate each participant's IPAQ score (informally, this represents the participants fitness level). We will be considering this as a candidate fixed effect in our model later.


```{r Nans}

# Helper to force numeric and turn NAs/blanks into 0
nz <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  x[is.na(x)] <- 0
  x
}

# Coerce all inputs
data$days_vigorous    <- nz(data$days_vigorous)
data$minutes_vigorous <- nz(data$minutes_vigorous)
data$days_moderate    <- nz(data$days_moderate)
data$minutes_moderate <- nz(data$minutes_moderate)
data$days_walking     <- nz(data$days_walking)
data$minutes_walking  <- nz(data$minutes_walking)   # important: was <chr>
data$hours_sitting    <- nz(data$hours_sitting)

# MET constants from IPAQ scoring guidelines
MET_VIG  <- 8.0
MET_MOD  <- 4.0
MET_WALK <- 3.3

# Component MET-min/week scores
data$vigorous_met <- ifelse(
  data$days_vigorous > 0 & data$minutes_vigorous > 0,
  data$days_vigorous * data$minutes_vigorous * MET_VIG,
  0
)

data$moderate_met <- ifelse(
  data$days_moderate > 0 & data$minutes_moderate > 0,
  data$days_moderate * data$minutes_moderate * MET_MOD,
  0
)

data$walking_met <- ifelse(
  data$days_walking > 0 & data$minutes_walking > 0,
  data$days_walking * data$minutes_walking * MET_WALK,
  0
)

# Total IPAQ MET-min/week
data$ipa_total_met <- data$vigorous_met + data$moderate_met + data$walking_met

# Number of days doing ANY activity (required for categorization)
data$total_days_any <- data$days_vigorous + data$days_moderate + data$days_walking

# IPAQ categorical classification HIGH / MODERATE / LOW
data$ipaq_category <- "Low"   # default

# High:
data$ipaq_category[
  (data$days_vigorous >= 3 & data$ipa_total_met >= 1500) |
  (data$total_days_any >= 7 & data$ipa_total_met >= 3000)
] <- "High"

# Moderate (only if not High):
data$ipaq_category[
  data$ipaq_category != "High" &
  (
    (data$days_vigorous >= 3 & data$minutes_vigorous >= 20) |
    (data$days_moderate >= 5 & data$minutes_moderate >= 30) |
    (data$days_walking  >= 5 & data$minutes_walking  >= 30) |
    (data$total_days_any >= 5 & data$ipa_total_met >= 600)
  )
] <- "Moderate"

# vigorous_met, moderate_met, walking_met, ipa_total_met, ipaq_category

# Optional quick checks:
#table(data$ipaq_category)
#summary(data$ipa_total_met)



```

## Selecting Data to analyze

We desire to analyze response times of "correct trials". We define "correct trials" as those in which the response received a score of 3.5 or higher on all four evaluation dimensions (accuracy, completeness, conciseness, and readability).
We used a conservative criterion by requiring responses to exceed the threshold on every quality dimension rather than allowing high performance on one dimension to compensate for poor performance on another. 

In addition, we filtered out trials in which participant reported not reading the AI output when given such output to consider.

For consistency, we also display the number of rows in our dataframe after selection.

```{r}
df_correct <- data |> 
  dplyr::filter( accuracy_both >= 3.5,
                completeness_both >= 3.5,
                conciseness_both >= 3.5,
                readability_both >= 3.5
                )


df_correct <- df_correct[df_correct$read_ai != "No",]

nrow(df_correct)

```




### Distribution of Response Times

For consistency and for reference, we examine the distribution of response time. Since each problem was designed to be completed in less than 900 seconds, we gain confidence if our mean response time is less than that.

Overall average response time
```{r}
mean_value <- mean((data$duration))
print(mean_value)
```


In this step, we plot our response times in a histogram. Since we view a skew, we transform our data to normalize it and satisfiy the assumptions of or mixed regression analysis. We employ a `sqrt` to skew.
```{r}
hist((df_correct$duration))
hist(sqrt(df_correct$duration))

```





## Mixed Effects Regression Analysis

### Specify Candidate Fixed Effects 

In this step, we first define the set of candidate fixed-effect terms to consider in the mixed-effects regression models. The core predictors (`exercise`, `AI`, and `sequence number`) are included in every model because they represent the primary experimental manipulations and account for trial order (i.e., potential learning or fatigue effects). 

Additional covariates and motivated interactions are then considered as optional terms during model comparison. These include `project_name` (controls for differences in difficulty across programming tasks), `ipaq_category` (accounts for differences in participants' physical activity levels), `occupation` (controls for variation associated with participants' professional background.), and `zoom_self` (which indicates whether participants completed the session by self-recording or under live supervision via Zoom).

```{r}

# Always include (core)
core_mains <- c("exercise", "ai", "seq_num")

# covariates
optional_mains <- c("project_name", "ipaq_category", "occupation", "zoom_self")

# Optional interactions
candidate_interactions <- c(
  "exercise * ai",
  "exercise * ipaq_category",
  "exercise * seq_num" # learning-curve moderation
)
```


In this step, we now generate a plausible fixed effects structures given that we want to keep 3 core fixed effects but would consider an extended model.
```{r}
make_fixed_forms <- function(core_mains, optional_mains, candidate_interactions) {
  fixed_forms <- character(0)
  
  ## 1) Parse interactions into list(name, vars)
  # e.g. "exercise * ai" -> name = "exercise:ai", vars = c("exercise","ai")
  interaction_defs <- lapply(candidate_interactions, function(trm) {
    vars <- trimws(strsplit(trm, "\\*")[[1]])
    list(
      name = paste(vars, collapse = ":"),  # use ":" in the formula
      vars = vars
    )
  })
  
  ## 2) All subsets of optional mains (including empty set)
  opt_subsets <- list(character(0))
  if (length(optional_mains) > 0) {
    for (k in 1:length(optional_mains)) {
      opt_subsets <- c(
        opt_subsets,
        combn(optional_mains, k, simplify = FALSE)
      )
    }
  }
  
  ## 3) All subsets of interactions (including empty)
  int_subsets <- list(integer(0))  # each element is a vector of indices into interaction_defs
  n_int <- length(interaction_defs)
  if (n_int > 0) {
    for (k in 1:n_int) {
      int_subsets <- c(
        int_subsets,
        combn(seq_len(n_int), k, simplify = FALSE)
      )
    }
  }
  
  ## 4) Build RHS strings
  for (opt_set in opt_subsets) {
    for (int_idx_set in int_subsets) {
      
      # which interactions are in this candidate?
      if (length(int_idx_set) == 0) {
        int_terms <- character(0)
        int_vars  <- character(0)
      } else {
        int_defs  <- interaction_defs[int_idx_set]
        int_terms <- vapply(int_defs, `[[`, character(1), "name")
        int_vars  <- unique(unlist(lapply(int_defs, `[[`, "vars")))
      }
      
      # all main-effect variables present in this model
      mains_here <- unique(c(core_mains, opt_set))
      
      # hierarchy rule: only keep models where all interaction vars are in mains
      if (!all(int_vars %in% mains_here)) {
        next
      }
      
      # assemble full RHS term list
      terms <- c(core_mains, opt_set, int_terms)
      rhs   <- paste(terms, collapse = " + ")
      fixed_forms <- c(fixed_forms, rhs)
    }
  }
  
  unique(fixed_forms)
}

fixed_forms <- make_fixed_forms(core_mains, optional_mains, candidate_interactions)
fixed_effects <- fixed_forms
#fixed_effects
```




### Specify candidate random effect structures for mixed model formulas

In this step, we next define a set of candidate random-effects structures for the mixed-effects regression models. These models account for the repeated-measures nature of the experiment by allowing observations from the same participant and programming task to be correlated. 

We consider models with random intercepts for participants (`1 | part_id`) and programming tasks (`1 | stim_src_id`), as well as models that allow the effects of AI (`ai`) and exercise (`exercise`) to vary across participants through random slopes.
To select an appropriate random-effects structure, we fit each candidate model using the core fixed effects (`exercise`, `ai`, and `seq_num`) and compare model fit using the Akaike Information Criterion (AIC).  BIC is also reported for transparency.


```{r}

random_structures <- c(
  R01 = "(1 | part_id)", 
  R02 = "(1 | stim_src_id)", 
  R0  = "(1 | part_id) + (1 | stim_src_id)",                 # intercepts only
  R1  = "(1 + ai | part_id) + (1 | stim_src_id)",            # AI slope
  R2  = "(1 + exercise | part_id) + (1 | stim_src_id)",      # exercise slope
  R3  = "(1 + ai + exercise | part_id) + (1 | stim_src_id)" # both slopes (correlated)
  #R3u = "(1 + ai + exercise || part_id) + (1 | stim_src_id)" # both slopes, no corr
)
```


Checking what random effect model is best given our core fixed effects and DV

```{r}


## 2. Main fixed-effects part of the model
fixed_main <- "sqrt(duration) ~ exercise + ai + seq_num"


## 4. Helper that fits all random structures and catches errors
fit_random_structures <- function(df, fixed_main, random_structures) {
  fits <- lapply(random_structures, function(re) {
    f_txt <- paste(fixed_main, "+", re)
    f <- as.formula(f_txt)
    message("\nFitting: ", f_txt)

    fit <- tryCatch(
      {
        lmer(f, data=df_correct, REML=FALSE)
      },
      error = function(e) {
        message("  ERROR: ", e$message)
        return(NULL)
      },
      warning = function(w) {
        message("  WARNING: ", w$message)
        invokeRestart("muffleWarning")  # show once, then silence
      }
    )

    fit
  })

  names(fits) <- random_structures
  fits
}

## 5. Run the fits

m_rand <- fit_random_structures(df, fixed_main, random_structures)

## 6. Drop models that failed (NULL)
m_rand_ok <- Filter(Negate(is.null), m_rand)

## 7. Compare AIC and BIC
aic_tab <- sapply(m_rand_ok, AIC)
bic_tab <- sapply(m_rand_ok, BIC)

#cat("\nAIC by random structure:\n")
#print(aic_tab)

#cat("\nBIC by random structure:\n")
#print(bic_tab)

## 8. Pick the best random structure by AIC (or BIC)
best_by_aic <- names(aic_tab)[which.min(aic_tab)]
best_by_bic <- names(aic_tab)[which.min(bic_tab)]
cat("\nBest random structure by AIC:", best_by_aic, "\n")
cat("\nBest random structure by BIC:", best_by_bic, "\n")

#best_model <- m_rand_ok[[best_by_aic]]

## 9. Inspect the chosen model

#summary(best_model)


```

Based on AIC, the selected random-effects structure includes random intercepts for both participant and programming task: `(1 | part_id) + (1 | stim_src_id)`.

```{r}

random_exprs = "(1 | part_id) + (1 | stim_src_id)"    #for reaction_time
  
```



Now, we combine the response variable, fixed effects, and random effects together to generate our candidate model structures. We consider 96 models.

```{r}
loglinear_formulae = c()
for (fe in fixed_effects) {
  for (re in random_exprs) {
    f = ""
    if (fe == "") {
      f = sprintf("sqrt(duration) ~ %s", re)
    } else {
      f = sprintf("sqrt(duration) ~ %s + %s", fe, re)
    }
    loglinear_formulae = append(loglinear_formulae, f)
  }
}

print(length(loglinear_formulae))
```


### Model fitting (\~2 minutes)

Here, we fit all the models. Of note, we use MLE rather than REML for compatibility with the AIC model selection strategy.

We also store BIC (an alternative model selection criterion) in addition to AIC here. The two have different theoretical properties, both of which may be useful. Both describe the log likelihood of a fitted model, with a penalty for model complexity. The penalty in BIC is higher than that for AIC, so optimal-BIC models tend to be smaller (with fewer parameters) than optimal-AIC models. We have done our analysis relying on AIC, but we provide BIC-based model selection here in case the reader is interested.

```{r eval=FALSE}
if (TRUE) {
  # make an output directory for the output of all the candidate linear models
  logtime_outputdir = 
    sprintf("%s/sqrt_time_model_correct_summaries_only_7_1_26", outputdir)
  mkdir(logtime_outputdir)
  
  # make an output file for the AIC and BIC of each model
  metricsfile = sprintf("%s/metrics.csv", logtime_outputdir)
  file.create(metricsfile)
  
  sink(file = metricsfile,
       append = FALSE,
       type = "output",
       split = FALSE
       )
  
  cat("formula,AIC,BIC\n")
  
  sink()
  
  for (f in loglinear_formulae) {
    # make an output directory for the output of this specific candidate model
    f_outputdir = sprintf("%s/%s", logtime_outputdir, gsub(" ", "", f, fixed = TRUE))
    mkdir(f_outputdir)
    
    # make the individual output file
    summaryfile = sprintf("%s/summary.txt", f_outputdir)
    
    file.create(summaryfile)
    
    # print a delimiter for console output
    cat(sprintf("\n---------------------------------------------\n%s\n", f))
    
    # fit the model
    m = lmer(f, data=df_correct, REML=FALSE)
    
    # send the model summary to the summary file
    sink(file = summaryfile,
         append = FALSE,
         type = "output",
         split = FALSE
         )
    
    print(summary(m))
    
    sink()
    
    # send the AIC and BIC to the metrics file
    metrics = sprintf("%s,%f,%f\n", f, AIC(m), BIC(m))
    
    sink(file = metricsfile,
         append = TRUE,
         type = "output",
         split = FALSE)
    
    cat(metrics)
    
    sink()
  }
}
```


### Model selection

```{r}
# read in the AIC/BIC data
fin = sprintf("%s/sqrt_time_model_correct_summaries_only_6_29_26/metrics.csv", outputdir)
log_time_models_da = read_csv(fin)
```
```{r}
head(log_time_models_da[order(log_time_models_da$AIC),], n = 3)
```
```{r}
head(log_time_models_da[order(log_time_models_da$BIC),], n = 3)
```



```{r}
fA = log_time_models_da[order(log_time_models_da$AIC),]$formula[[1]]
fB = log_time_models_da[order(log_time_models_da$BIC),]$formula[[1]]
```

We note that it is useful that AIC and BIC agree on the top ranked model (#1). When AIC and BIC agree on the highest-ranked model, we treat this as additional evidence that the selected model provides a robust balance between fit and complexity. In such cases, we retain the top-ranked model for subsequent analyses.


```{r}
m0 = lmer(fA, data=df_correct, REML=FALSE)
summary(m0)
```


## Research Question 1 (Response time during the exericse treatment)

We next evaluate the statistical significance of each fixed effect using likelihood ratio tests. Each test compares the selected model against a nested model that excludes the predictor of interest.

```{r}
fC=  sqrt(duration) ~ ai + seq_num + (1 | part_id) + (1 | stim_src_id)
fD = sqrt(duration) ~ exercise + seq_num +   (1 | part_id) + (1 | stim_src_id)
fE = sqrt(duration) ~ exercise + ai +   (1 | part_id) + (1 | stim_src_id)

mC = lmer(fC, data=df_correct, REML=FALSE)
mD = lmer(fD, data=df_correct, REML=FALSE)
mE = lmer(fE, data=df_correct, REML=FALSE)

```

```{r}

anova(m0, mC)  # Test contribution of exercise
anova(m0, mD)  # Test contribution of AI
anova(m0, mE)  # Test contribution of seq_num (learning effects)

```

To quantify the overall explanatory power of the mixed-effects model, we report both the marginal and conditional (R^2) values. In mixed-effects models, a single (R^2) is not sufficient because variance is explained by both the fixed effects (the experimental predictors) and the random effects (differences among participants and stimuli). The marginal (R^2) represents the proportion of variance explained by the fixed effects alone, whereas the conditional (R^2) represents the proportion of variance explained by the full model, including both fixed and random effects. Reporting both measures provides a more complete assessment of model fit and the relative contributions of the experimental manipulations and the hierarchical structure of the data.

```{r}
summary(m0)
r2(m0)
```





